---
title: "Banking Insurance Product – Tree-Based Models"
output: html_document
---

## 1. Setup and Data

```{r}
## Libraries
# library(tidyverse)
# library(DescTools)
# library(dplyr)
# library(tidyr)
# library(caret)
# library(Metrics)
# library(ROCit)
# library(car)
# library(randomForest)
# library(pROC)
# library(rpart)
# library(rpart.plot)
# library(xgboost)
# library(Ckmeans.1d.dp)
# library(pdp)

theme_set(theme_minimal())

# set seed
set.seed(4321)


# Load training/validation sets
ins_t <- read.csv('insurance_t.csv')
```

```{r}
# Checking missing values 
colSums(is.na(ins_t))
```

## 2. Data Preparation and Imputation

'Bank' standard: median (continuous), mode (categorical indicators).
Keep preprocessing minimal — no variable selection before modeling.

```{r}
# Making copy for imputations
train2 <- ins_t

# Mode function
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

# CONTINUOUS variables - use MEDIAN
continuous_impute <- c("ACCTAGE", "PHONE", "POSAMT", 
                       "INVBAL", "CCBAL", "CCPURC", 
                       "INCOME", "LORES", "HMVAL", "AGE", "CRSCORE", "POS")

for (var in continuous_impute) {
  if (any(is.na(train2[[var]]))) {
    train2[[var]][is.na(train2[[var]])] <- median(train2[[var]], na.rm = TRUE)
  }
}

# CATEGORICAL (indicators) - use MODE  
categorical_impute <- c("INV", "CC")

for (var in categorical_impute) {
  if (any(is.na(train2[[var]]))) {
    train2[[var]][is.na(train2[[var]])] <- get_mode(train2[[var]])
  }
}

# Store values for validation set - future hw
# median_vals <- sapply(train2[continuous_impute], median, na.rm = TRUE)
# mode_vals <- sapply(train2[categorical_impute], get_mode)

# removing HMOWN NA's
train2 <- na.omit(train2)

# Verify no more nulls
colSums(is.na(train2))

imputed_train <- train2



```

## 3. Random Forest Model

### 3.1 Convert Target to Factor (Required for Classification)

```{r}
# Convert INS to factor for classification
train2$INS <- as.factor(train2$INS)

# Convert to data frame (required for randomForest)
train <- as.data.frame(train2)

# Check the target variable
table(train$INS)

# Prop Table
prop.table(table(train$INS))
```

### 3.2 Initial Random Forest Model

Initial unbalanced and unedited model

```{r}
set.seed(12345)
rf.insurance.unbalanced <- randomForest(INS ~ ., 
                                        data = train, 
                                        ntree = 500, 
                                        nodesize = 1,
                                        importance = TRUE)
# View model summary
print(rf.insurance.unbalanced)

# Plot error vs number of trees
plot(rf.insurance.unbalanced, main = "Random Forest: Error vs Number of Trees")
legend("topright", 
       legend = c("OOB Error", "Class 0", "Class 1"),
       col = 1:3, 
       lty = 1:3,
       cex = 0.8)
```

First set of edits

```{r}
# Check class distribution
cat("Class distribution:\n")
print(table(train$INS))
print(prop.table(table(train$INS)))

# Build next random forest with class balancing
# Using sampsize to balance the classes in each bootstrap sample
minority_class_size <- min(table(train$INS))
```

```{r}
# Build next random forest with default parameters; edited sample size and ntree
set.seed(12345)
rf.full <- randomForest(INS ~ ., 
                             data = train, 
                             ntree = 250, 
                             importance = TRUE,
                              nodesize = 1,
                             sampsize = c(minority_class_size, minority_class_size))

# View model summary
print(rf.full)

# Plot error vs number of trees
plot(rf.full, main = "Random Forest: Error vs Number of Trees (Balanced)")
legend("topright", 
       legend = c("OOB Error", "Class 0", "Class 1"),
       col = 1:3, 
       lty = 1:3,
       cex = 0.8)
```

```{r}
# Compare with unbalanced model
cat("\n=== Comparison: Unbalanced vs Balanced ===\n")
cat("Unbalanced model:\n")
print(rf.insurance.unbalanced$confusion)
cat("\nBalanced model:\n")
print(rf.full$confusion)
```

### 3.3 Variable Selection Attempt Using Random Variable

```{r}
# Add random variable for variable selection sanity check
train$random <- rnorm(nrow(train))

# Rebuild with random variable
set.seed(12345)
rf.insurance.random <- randomForest(INS ~ ., 
                                    data = train, 
                                    ntree = 500, 
                                    importance = TRUE)

# Plot variable importance
varImpPlot(rf.insurance.random, 
           sort = TRUE, 
           n.var = min(20, ncol(train)-1),
           main = "Variable Importance - Check for Variables Below Random")

# Get importance values
importance_df <- as.data.frame(importance(rf.insurance.random))
importance_df$Variable <- rownames(importance_df)
importance_df <- importance_df[order(-importance_df$MeanDecreaseGini),]

# Show all variables with importance
print(importance_df)

# Identify variables below random variable
random_importance <- importance_df[importance_df$Variable == "random", "MeanDecreaseGini"]
cat("\nRandom variable importance (MeanDecreaseGini):", random_importance, "\n")

# Variables to potentially remove (below or near random)
weak_vars <- importance_df[importance_df$MeanDecreaseGini <= random_importance * 1.2, "Variable"]
cat("\nVariables below or near random variable:\n")
print(weak_vars)
```

Remove random variable after

```{r}
train$random <- NULL
importance_df <- subset(importance_df, Variable != "random")
```

### 3.3.5 Variable Selection Using Cross-Validation Performance Approach

```{r}
# Extract and rank variables by importance
importance_full <- as.data.frame(importance(rf.full))
importance_full$Variable <- rownames(importance_full)
importance_full <- importance_full[order(-importance_full$MeanDecreaseGini),]

cat("\n=== Variable Importance Rankings ===\n")
print(importance_full[, c("Variable", "MeanDecreaseGini")])
```

```{r}
# Plot full importance
varImpPlot(rf.full, 
           sort = TRUE,
           main = "Variable Importance - Full Model")
```

### 3.4 OOB Testing

#### Testing different numbers of variables using OOB Error

Finding minimum OOB for current random forest parameters Test models with top 5, 10, 15, 20, 25, 30, 35, all variables

```{r}
n_vars_to_test <- c(5, 10, 15, 20, 25, 30, 35, 37)

# empty data frame for results
oob_results <- data.frame( 
  n_variables = integer(),
  oob_error = numeric(),
  class_0_error = numeric(),
  class_1_error = numeric()
  )


set.seed(12345)
for (n_vars in n_vars_to_test) {
  selected_vars <- head(importance_full$Variable, n_vars)
  train_subset <- train[, c(selected_vars, "INS")]
  train_subset$INS <- as.factor(train_subset$INS)
  
  rf_model <- randomForest(
    INS ~ .,
    data = train_subset,
    ntree = 250,
    importance = TRUE,
    nodesize = 1,
    sampsize = c(minority_class_size, minority_class_size)
  )
  
  # Extract OOB + class-specific errors
  last_row <- rf_model$err.rate[nrow(rf_model$err.rate), ]
  oob_error <- last_row["OOB"]
  class_0_error <- last_row["0"]
  class_1_error <- last_row["1"]
  
  oob_results <- rbind(oob_results, data.frame(
    n_variables = n_vars,
    oob_error = oob_error,
    class_0_error = class_0_error,
    class_1_error = class_1_error
  ))
}

# Check results table
print(oob_results)
```

#### Plot OOB Error vs Number of Variables

```{r}
# Plot with ggplot2
ggplot(oob_results, aes(x = n_variables, y = oob_error)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "darkred", size = 3) +
  theme_minimal(base_size = 13) +
  labs(
    title = "OOB Error vs. Number of Variables",
    x = "Number of Variables",
    y = "OOB Error Rate"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

#### Plot class-specific errors too

```{r}
library(reshape2)

# Reshape data for plotting multiple lines
oob_long <- melt(
  oob_results,
  id.vars = "n_variables",
  variable.name = "Error_Type",
  value.name = "Error_Rate"
)

ggplot(oob_long, aes(x = n_variables, y = Error_Rate, color = Error_Type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  theme_minimal(base_size = 13) +
  labs(
    title = "OOB Error by Number of Variables",
    x = "Number of Variables",
    y = "Error Rate",
    color = "Error Type"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

### 3.5 Select variables based on CV results and interpretation

```{r}
# Rule of thumb: Accept models within 1-2% of minimum error for parsimony
tolerance <- 0.02  # 2% tolerance
acceptable_models <- oob_results[oob_results$oob_error <= min(oob_results$oob_error) * (1 + tolerance), ]

cat("\nModels within", tolerance*100, "% of minimum error:\n")
print(acceptable_models[, c("n_variables", "oob_error")])
```

Models with 25, 30, and 37 fit the OOB error threshold

#### Test 25 variables for parsimony

```{r}
# Choose simplest acceptable model
simplest_n_vars <- min(acceptable_models$n_variables)
cat("\n=== SIMPLEST SELECTION ===\n")
cat("Selected number of variables: 25 \n")
cat("OOB Error:", round(acceptable_models$oob_error[acceptable_models$n_variables == 25], 4), "\n")
cat("vs. Absolute minimum:", round(min(oob_results$oob_error), 4), "\n")

# Create reduced selected dataset
simplest_selected_vars <- head(importance_full$Variable, simplest_n_vars)
cat("\nVariables in reduced model:\n")
print(simplest_selected_vars)

train.selected <- train[, c(simplest_selected_vars, "INS")]
train.selected$INS <- as.factor(train.selected$INS)
```

### 3.6 Visualizing and testing selected variables

#### 25 Variables - visualizing the selected vs non-selected variables

```{r}
importance_full$Selected <- ifelse(importance_full$Variable %in% simplest_selected_vars, 
                                  "Kept", "Removed")

ggplot(importance_full, aes(x = reorder(Variable, MeanDecreaseGini), 
                           y = MeanDecreaseGini,
                           fill = Selected)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Kept" = "steelblue", "Removed" = "gray70")) +
  coord_flip() +
  labs(title = paste("Variable Selection: 25 Variables Kept"),
       subtitle = "Based on cross-validation performance",
       x = "Variable",
       y = "Mean Decrease Gini (Importance)") +
  theme_minimal() +
  theme(legend.position = "bottom")

cat("\n=== Variable Selection Summary ===\n")
cat("Original variables:", ncol(train) - 1, "\n")
cat("Variables selected: 25 \n")
cat("Variables removed:", (ncol(train) - 1) - 25, "\n")
cat("\nSelection criterion: Minimize OOB error with parsimony\n")
```

#### Test 30 vars for parsimony

```{r}
# Choose middle simple acceptable model
simple_n_vars <- 30
cat("\n=== SIMPLEST SELECTION ===\n")
cat("Selected number of variables: 30 \n")

# Create reduced selected dataset
simple_selected_vars <- head(importance_full$Variable, simple_n_vars)
cat("\nVariables in reduced model:\n")
print(simple_selected_vars)

train.selected <- train[, c(simple_selected_vars, "INS")]
train.selected$INS <- as.factor(train.selected$INS)
```

Visualize the selected vs non-selected variables

```{r}
importance_full$simple_selected_vars <- ifelse(importance_full$Variable %in% simple_selected_vars, "Kept", "Removed")

ggplot(importance_full, aes(x = reorder(Variable, MeanDecreaseGini), 
                           y = MeanDecreaseGini,
                           fill = simple_selected_vars)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Kept" = "steelblue", "Removed" = "gray70")) +
  coord_flip() +
  labs(title = paste("Variable Selection: 30 Variables Kept"),
       subtitle = "Based on cross-validation performance",
       x = "Variable",
       y = "Mean Decrease Gini (Importance)") +
  theme_minimal() +
  theme(legend.position = "bottom")

cat("\n=== Variable Selection Summary ===\n")
cat("Original variables:", ncol(train) - 1, "\n")
cat("Variables selected: 30 \n")
cat("Variables removed:", (ncol(train) - 1) - 30, "\n")
cat("\nSelection criterion: Minimize OOB error with parsimony\n")
```

### 3.7 Tested tuned mtry Parameter (gives default which is 5)

```{r}
# Tune mtry (number of variables considered at each split)
# Default for classification is sqrt(p)
set.seed(12345)

cat("\n=== Tuning mtry Parameter ===\n")
cat("Default mtry for classification: sqrt(p) =", floor(sqrt(ncol(train.selected) - 1)), "\n\n")

tuneRF_result <- tuneRF(x = train.selected[, -which(names(train.selected) == "INS")], 
                        y = train.selected$INS,
                        plot = TRUE, 
                        ntreeTry = 250, 
                        stepFactor = 0.5,
                        improve = 0.01,
                        trace = TRUE,
                        doBest = FALSE,
                        nodesize = 1,
                        sampsize = c(minority_class_size, minority_class_size))

print(tuneRF_result)

# Get optimal mtry
optimal_mtry <- tuneRF_result[which.min(tuneRF_result[,2]), 1]
cat("\nOptimal mtry:", optimal_mtry, "\n")
cat("Default mtry (sqrt(p)):", floor(sqrt(ncol(train.selected) - 1)), "\n")
```

### 3.8 Reduced Variable Random Forest Model

#### 25 variables

```{r}
# Build reduced varaible model with selected variables, nodesize, and balanced sampling
set.seed(12345)
rf.insurance.reduced <- randomForest(INS ~ ., 
                                   data = train.selected, 
                                   ntree = 250,
                                   nodesize = 1,
                                   importance = TRUE,
                                   sampsize = c(minority_class_size, minority_class_size))

cat("\n=== Reduced Var Random Forest Model ===\n")
print(rf.insurance.reduced)

# Plot final model error
plot(rf.insurance.reduced, main = "Reduced Var Random Forest: Error vs Trees")
legend("topright", 
       legend = c("OOB Error", "Class 0", "Class 1"),
       col = 1:3, 
       lty = 1:3,
       cex = 0.8)
```

#### 30 Variables

```{r}
selected_vars <- head(importance_full$Variable, 30)
train_subset <- train[, c(selected_vars, "INS")]
train_subset$INS <- as.factor(train_subset$INS)
  
rf.mild.reduced <- randomForest(
    INS ~ .,
    data = train_subset,
    ntree = 250,
    importance = TRUE,
    nodesize = 1,
    sampsize = c(minority_class_size, minority_class_size)
  )

print(rf.mild.reduced)
```

### 3.9 Comparing to full balanced model from earlier

```{r}
set.seed(12345)
rf.full.final <- randomForest(INS ~ ., 
                             data = train, 
                             ntree = 250, 
                             importance = TRUE,
                             nodesize = 1,
                             sampsize = c(minority_class_size, minority_class_size))

print(rf.full.final)
```

```{r}
# Compare all models
cat("\n=== Model Performance Comparison ===\n")
cat("Unbalanced Model (37 vars):\n")
cat("  OOB Error:", round(mean(rf.insurance.unbalanced$err.rate[,1]), 4), "\n")
cat("  Class 0 Error:", round(rf.insurance.unbalanced$confusion[1, "class.error"], 4), "\n")
cat("  Class 1 Error:", round(rf.insurance.unbalanced$confusion[2, "class.error"], 4), "\n\n")

cat("Reduced Var Tuned Model (25 vars):\n")
cat("  OOB Error:", round(mean(rf.insurance.reduced$err.rate[,1]), 4), "\n")
cat("  Class 0 Error:", round(rf.insurance.reduced$confusion[1, "class.error"], 4), "\n")
cat("  Class 1 Error:", round(rf.insurance.reduced$confusion[2, "class.error"], 4), "\n\n")

cat("Mildly Reduced Var Tuned Model (30 vars):\n")
cat("  OOB Error:", round(mean(rf.mild.reduced$err.rate[,1]), 4), "\n")
cat("  Class 0 Error:", round(rf.mild.reduced$confusion[1, "class.error"], 4), "\n")
cat("  Class 1 Error:", round(rf.mild.reduced$confusion[2, "class.error"], 4), "\n\n")

cat("Final Full Balanced Model (37 vars):\n")
cat("  OOB Error:", round(mean(rf.full.final$err.rate[,1]), 4), "\n")
cat("  Class 0 Error:", round(rf.full.final$confusion[1, "class.error"], 4), "\n")
cat("  Class 1 Error:", round(rf.full.final$confusion[2, "class.error"], 4), "\n\n")

```

### 3.95 Testing Accuracy and AUC between 30 and 37 models

#### 30 var (rf.mild.reduced)

```{r}
# Get OOB predictions (built-in cross-validation from Random Forest)
# Each observation is predicted by trees that didn't use it in training
oob_pred_class <- predict(rf.mild.reduced, type = "response")
oob_pred_prob <- predict(rf.mild.reduced, type = "prob")[,2]

# Confusion matrix
conf_matrix <- table(Predicted = oob_pred_class, Actual = train.selected$INS)
cat("\n=== Confusion Matrix (Out-of-Bag) ===\n")
print(conf_matrix)

# Calculate accuracy
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
cat("\nAccuracy (OOB):", round(accuracy, 4), "\n")

# use ROCit package for consistency with logistic regression
roc_empirical <- rocit(score = oob_pred_prob, 
                       class = train.selected$INS)

plot(roc_empirical, 
     col = c("blue", "red"), 
     legend = FALSE,
     YIndex = FALSE,
     main = "ROC Curve - Random Forest (ROCit)")
grid()

# Get AUC from ROCit
auc_rocit <- roc_empirical$AUC
cat("AUC (ROCit):", round(auc_rocit, 4), "\n")

# Add AUC text to plot
text(0.6, 0.2, paste("AUC =", round(auc_rocit, 4)), cex = 1.3, font = 2)

```

#### 37 var (rf.full.final)

```{r}
# Get OOB predictions (built-in cross-validation from Random Forest)
# Each observation is predicted by trees that didn't use it in training
oob_pred_class <- predict(rf.full.final, type = "response")
oob_pred_prob <- predict(rf.full.final, type = "prob")[,2]

# Confusion matrix
conf_matrix <- table(Predicted = oob_pred_class, Actual = train.selected$INS)
cat("\n=== Confusion Matrix (Out-of-Bag) ===\n")
print(conf_matrix)

# Calculate accuracy
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
cat("\nAccuracy (OOB):", round(accuracy, 4), "\n")

# use ROCit package for consistency with logistic regression
roc_empirical <- rocit(score = oob_pred_prob, 
                       class = train.selected$INS)

plot(roc_empirical, 
     col = c("blue", "red"), 
     legend = FALSE,
     YIndex = FALSE,
     main = "ROC Curve - Random Forest (ROCit)")
grid()

# Get AUC from ROCit
auc_rocit <- roc_empirical$AUC
cat("AUC (ROCit):", round(auc_rocit, 4), "\n")

# Add AUC text to plot
text(0.6, 0.2, paste("AUC =", round(auc_rocit, 4)), cex = 1.3, font = 2)

```

FINAL MODEL CHOSEN FOR PARSIMONY - BALANCED MODEL WITH 30 VAR

### Variable Importance for Final Model

```{r}
# Plot variable importance for final model
varImpPlot(rf.mild.reduced, 
           sort = TRUE,
           main = paste("Final Model: Variable Importance (30 vars):\n"))

# Get importance values - using MeanDecreaseGini as the ranking metric
importance_final <- as.data.frame(importance(rf.mild.reduced))
importance_final$Variable <- rownames(importance_final)
importance_final <- importance_final[order(-importance_final$MeanDecreaseGini),]

# Display variable importance ranked by MeanDecreaseGini
cat("\n=== Variable Importance (Ranked by MeanDecreaseGini) ===\n")
print(importance_final[, c("Variable", "MeanDecreaseGini")])

# Create a clean visualization
ggplot(importance_final, aes(x = reorder(Variable, MeanDecreaseGini), 
                             y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance - Final Random Forest Model",
       x = "Variable",
       y = "Mean Decrease Gini") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 9))
```

## 4. Model Evaluation Using Out-of-Bag Predictions

### 4.1 OOB Predictions and Performance

```{r}
# Get OOB predictions (built-in cross-validation from Random Forest)
# Each observation is predicted by trees that didn't use it in training
oob_pred_class <- predict(rf.mild.reduced, type = "response")
oob_pred_prob <- predict(rf.mild.reduced, type = "prob")[,2]

# Confusion matrix
conf_matrix <- table(Predicted = oob_pred_class, Actual = train.selected$INS)
cat("\n=== Confusion Matrix (Out-of-Bag) ===\n")
print(conf_matrix)

# Calculate accuracy
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
cat("\nAccuracy (OOB):", round(accuracy, 4), "\n")
```

### 4.2 ROC Curve and AUC

```{r}
# Create ROC object using ROCit for consistency
roc_rf_rocit <- rocit(score = oob_pred_prob, 
                      class = train.selected$INS)

# Print AUC
print("Area Under ROC Curve (AUC):")
print(round(roc_rf_rocit$AUC, 4))

# Plot ROC curve
plot(roc_rf_rocit, col = "blue", lwd = 2)
title(main = "ROC Curve for Random Forest (Training Data)")
grid()

# Add AUC text to plot
text(0.6, 0.3, paste("AUC =", round(auc_rocit, 4)), cex = 1.3, font = 2)
```

### 4.3 Additional Performance Metrics

```{r}
# Calculate additional metrics
TP <- conf_matrix[2,2]
TN <- conf_matrix[1,1]
FP <- conf_matrix[2,1]
FN <- conf_matrix[1,2]

sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)
precision <- TP / (TP + FP)
f1_score <- 2 * (precision * sensitivity) / (precision + sensitivity)

cat("\n=== Performance Metrics (Out-of-Bag) ===\n")
cat("Sensitivity (Recall):", round(sensitivity, 4), "\n")
cat("Specificity:", round(specificity, 4), "\n")
cat("Precision:", round(precision, 4), "\n")
cat("F1 Score:", round(f1_score, 4), "\n")

# Create comprehensive summary table
performance_summary <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "Precision", "F1 Score", "AUC"),
  Value = round(c(accuracy, sensitivity, specificity, precision, f1_score, auc_rocit), 4)
)
cat("\n=== Performance Summary Table ===\n")
print(performance_summary, row.names = FALSE)
```

#### Checking R\^2

```{r}
# Using predicted probabilities (continuous values 0-1)
# Treat actual classes as numeric (0 and 1) for these calculations
actual_numeric <- as.numeric(train.selected$INS) - 1  # Convert factor to 0/1
pred_prob <- oob_pred_prob  # Already have this from earlier

# R² (treating as regression problem)
# R² = 1 - (SS_residual / SS_total)
ss_residual <- sum((actual_numeric - pred_prob)^2)
ss_total <- sum((actual_numeric - mean(actual_numeric))^2)
r_squared <- 1 - (ss_residual / ss_total)

cat("R² (Pseudo R-squared):", round(r_squared, 4), "\n")
cat("  Interpretation: Model explains", round(r_squared * 100, 1), 
    "% of variance in purchase behavior\n")
```

## 5. XGBoost Model

### 5.1 Initial Model and Interpretation

```{r}
#Create random variable
imputed_train$random <- rnorm(nrow(imputed_train))

#Split X and y
#model.matrix uses OHE
train_x <- model.matrix(INS ~ ., data = imputed_train)[, -1]
train_y <- imputed_train$INS
```

```{r}
set.seed(4321)

#Build XGB- set objective = binary:logistic for classification problem
#subsample = 0.5 prevents overfitting
#nrounds = 100 is the number of boosting iterations (ie trees)
xgb.ins_t <- xgboost(data = train_x, 
                     label = train_y, 
                     subsample = 0.5, 
                     nrounds = 100, 
                     objective = "binary:logistic", 
                     verbose = 0)
```

```{r}
set.seed(4321)

#Use cross validation
xgbcv.ins_t<- xgb.cv(
  data = train_x, 
  label = train_y, 
  subsample = 0.5, 
  nrounds = 100,
  nfold = 10, 
  objective = "binary:logistic", 
  verbose = 0
)

#Evaluation metrics for each boosting round and fold
print(xgbcv.ins_t)
```

```{r}
set.seed(4321)

tune_grid <- expand.grid(
  nrounds = 24,
  eta = c(0.1, 0.15, 0.2, 0.25, 0.3),
  max_depth = c(1:4),
  gamma = c(0,1,3),
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = c(0.25, 0.5, 0.75, 1)
)

```

```{r}
set.seed(4321)

#Change y to a factor for caret model
train_y <- factor(imputed_train$INS, levels = c(0,1), labels = c("No","Yes"))

xgb.caret <- train(
  x = train_x, 
  y = train_y,
  method = "xgbTree",
  tuneGrid = tune_grid,
  trControl = trainControl(method = 'cv', number = 10, classProbs = TRUE, summaryFunction = twoClassSummary),
  metric = 'ROC'
)
```

```{r}
plot(xgb.caret)
xgb.caret$bestTune
```

### 5.2 Tuning Model and Variable Importance

```{r}
train_y <- as.numeric(imputed_train$INS)   # ensure 0/1 numeric

# Fit tuned model with best shrinkage (eta), depth (3)
xgb.final <- xgboost(
  data = train_x, 
  label = train_y, 
  subsample = 1,
  min_child_weight = 5,
  nrounds = 24, 
  gamma = 3,
  eta = 0.25, 
  max_depth = 4,
  objective = "binary:logistic",
  verbose = 0
)
```

```{r}
# Variable importance including random variable
xgb.ggplot.importance(xgb.importance(feature_names = colnames(train_x), model = xgb.final))

imp_table <- xgb.importance(
  feature_names = colnames(train_x),
  model = xgb.final
)

# View as a table
print(imp_table)
```

```{r}
#Check AUC on train
pred_train <- predict(xgb.final, train_x)

roc_train <- pROC::roc(response = train_y,
                       predictor = pred_train,
                       quiet = TRUE)

pROC::auc(roc_train)
```

### 5.3 ROC Curve and Additional Performance Metrics

```{r}
roc <- rocit(pred_train, train_y)

# Plot ROC with blue line + title
plot(roc,
     col = "blue",
     lwd = 2,
     YI = TRUE,
     diag = TRUE)

title("ROC Curve for XGBoost Model (Training Data)", cex.main = 1)

auc_val <- round(roc$AUC, 3)
text(
  0.65, 0.4,
  labels = paste0("AUC = ", auc_val),
  cex = 1,
  col = "black",
  font = 2       
)

summary(roc)
```

```{r}
#Confusion Matrix with optimal cutoff
cutoff <- 0.3441080
pred_class <- ifelse(pred_train > cutoff, 1, 0)
table(Predicted = pred_class, Actual = train_y)
```

```{r}
#View Correlation Matrix
pred_factor <- factor(pred_class, levels = c(0,1))
actual_factor <- factor(train_y, levels = c(0,1))

cm <- confusionMatrix(pred_factor, actual_factor, positive = "1")
cm
```

```{r}
accuracy     <- cm$overall["Accuracy"]
sensitivity  <- cm$byClass["Sensitivity"]
specificity  <- cm$byClass["Specificity"]
precision    <- cm$byClass["Pos Pred Value"]  
f1           <- cm$byClass["F1"]

round(c(accuracy, sensitivity, specificity, precision, f1), 4)
```
